---
source: Rmd
title: "Synapse and AD Knowledge Portal"
teaching: 40
exercises: 10
questions:
- "How to work with Synapse R client?"
- "How to work with data in AD Knowledge Portal?"
objectives:
- "Explain how to use Synapser Package."
- "Demonstrate how to locate data and metadata in the Portal."
- "Demonstrate how to download data from the Portal programmatically."
keypoints:
- "Use your Synapse login credentials to access the Portal."
- "Use Synapser package to download data from the Portal."
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

Author: Laura Heath / Sage Bionetworks


# Downloading and Exploring Data from Synapse

Before you begin, determine your working directory (get(wd) and set(wd)), and create two new folders within your preferred working directory entitled "data" and "results"

## Install and Load Packages

If you haven't already, install synapser (the Synapse R client), as well
as the tidyverse family of packages
```{r}
# install synapser
install.packages("synapser", repos = c("http://ran.synapse.org", "http://cran.fhcrc.org"))

# install tidyverse if you don't already have it
install.packages("tidyverse")
```
Load libraries
```{r}
library(synapser)
library(tidyverse)
#library(clusterProfiler)
```
## Login to Synapse

Next, you will need to log in to your Synapse account.

*Login option 1*: Synapser takes credentials from your Synapse web session
If you are logged into the Synapse web browser, synapser will
automatically use your login credentials to log you in during your R
session! All you have to do is:

```{r}
#synLogin()
```

If for whatever reason that didn’t work, try one of these options:

*Login option 2*: Synapse username and password In the code below, replace
the \<\> with your Synapse username and password.

```{r}
#synLogin("<username>", "<password>")
```

*Login option 3*: Synapse PAT If you usually log in to Synapse with your
Google account, you will need to use a Synapser Personal Access Token
(PAT) to log in with the R client. Follow these instructions to generate
a personal access token, then paste the PAT into the code below. Make
sure you scope your access token to allow you to View, Download, and
Modify.

```{r}
#synLogin(authToken = "<paste your personal access token here>")
```

```{r, echo = FALSE}
synLogin(authToken = "eyJ0eXAiOiJKV1QiLCJraWQiOiJXN05OOldMSlQ6SjVSSzpMN1RMOlQ3TDc6M1ZYNjpKRU9VOjY0NFI6VTNJWDo1S1oyOjdaQ0s6RlBUSCIsImFsZyI6IlJTMjU2In0.eyJhY2Nlc3MiOnsic2NvcGUiOlsidmlldyIsImRvd25sb2FkIiwibW9kaWZ5Il0sIm9pZGNfY2xhaW1zIjp7fX0sInRva2VuX3R5cGUiOiJQRVJTT05BTF9BQ0NFU1NfVE9LRU4iLCJpc3MiOiJodHRwczovL3JlcG8tcHJvZC5wcm9kLnNhZ2ViYXNlLm9yZy9hdXRoL3YxIiwiYXVkIjoiMCIsIm5iZiI6MTcyMTc4OTUyNywiaWF0IjoxNzIxNzg5NTI3LCJqdGkiOiIxMDMxNSIsInN1YiI6IjMzNDc1MjgifQ.sr1euDKwv-U8PPj36ej7Kfo_Fne9KZOzu-fWZ2qeka8SsLwAs26z6623QyVUGHK2RoF_LeZ479d465BXxW3z47kbUUAJguzTVLlrB7pRHrvl7fFEAEq0fC6pQzv-WYYHuys7mB6HfAICJXpL2zHWA32OjKUjfI3_eIRZw9yyBFmrDHo8OSLH3PED6E_CNs2Ydhl_V8F9lsN5UTOLWyc6Rz-UR--JdzW7nLFGlKScwb0YxJyw15K0MMjNg8yBxZCxPCsD18AWGZNh4nI7ppod7wMaJG3J1akujRqNqYPJftPKN7BT_0hpgISsY3I8djbFzMVy7MlfUIRgkPfmp68GBQ")
```
For more information on managing Synapse credentials with synapser, see
the documentation here: https://r-docs.synapse.org/articles/manageSynapseCredentials.html.


## Download Data

While you can always download data from the AD Portal website via your web browser, it’s usually faster and often more convenient to download data programmatically.

**Download a single file** 
To download a single file from the AD Knowledge
Portal, you can click the linked file name to go to a page in the
Synapse platform where that file is stored. Using the synID on that
page, you can call the synGet() function from synapser to download the
file.

**Exercise 1**: Use Explore Data to find processed RNAseq data from the
Jax.IU.Pitt_5XFAD Study. 

https://adknowledgeportal.synapse.org/


This filters the table to a single file. In the
“Id” column for this htseqcounts_5XFAD.txt file, there is a unique
Synapse ID (synID).

We can then use that synID to download the file.
```{r}
counts_id <- "syn22108847"
synGet(counts_id, downloadLocation = "../data/")
```

## Bulk download files

**Exercise 2**: Use Explore Studies to find all metadata
files from the Jax.IU.Pitt_5XFAD study.

Use the facets and search bar to
look for data you want to download from the AD Knowledge Portal. Once
you’ve identified the files you want, click on the download arrow icon
on the top right of the Explore Data table and select “Programmatic
Options” from the drop-down menu.

In the window that pops up, select the “R” tab from the top menu bar.
This will display some R code that constructs a SQL query of the Synapse
data table that drives the AD Knowledge Portal. This query will allow us
to download only the files that meet our search criteria.

The function synTableQuery() returns a Synapse object wrapper around a
CSV file that is automatically downloaded to a Synapse cache directory
.synapseCache in your home directory. You can use query\$filepath to see
the path to the file in the Synapse cache.

```{r}
# download the results of the filtered table query
query <- synTableQuery("SELECT * FROM syn11346063.52 WHERE ( ( `study` HAS ( 'Jax.IU.Pitt_5XFAD' ) ) AND ( `resourceType` = 'metadata' ) )")

# view the file path of the resulting csv
query$filepath
```

We’ll use read.csv to read the CSV file into R (although the provided
read.table or any other base R version is also fine!). We can explore
the download_table object and see that it contains information on all of
the AD Portal data files we want to download. Some columns like the “id”
and “parentId” columns contain info about where the file is in Synapse,
and some columns contain AD Portal annotations for each file, like
“dataType”, “specimenID”, and “assay”. This annotation table will later
allow us to link downloaded files to additional metadata variables!
```{r}
# read in the table query csv file
download_table <- read_csv(query$filepath, show_col_types = FALSE)

download_table
```

Finally, we use a mapping function from the purrr package to loop
through the “id” column and apply the synGet() function to each file’s
synID. In this case, we use purrr::walk() because it lets us call
synGet() for its side effect (downloading files to a location we
specify), and returns nothing.
```{r}
# loop through the column of synIDs and download each file
purrr::walk(download_table$id, ~synGet(.x, downloadLocation = "../data/"))
```

Congratulations, you have bulk downloaded files from the AD Knowledge
Portal!

**An important note**: for situations where you are downloading many large
files, the R client performs substantially slower than the command line
client or the Python client. In these cases, you can use the
instructions and code snippets for the command line or Python client
provided in the “Programmatic Options” menu.


## Working with AD Portal metadata 

**Metadata basics** 

We have now downloaded
several metadata files and an RNAseq counts file from the portal. For
our next exercises, we want to read those files in as R data so we can
work with them.

We can see from the download_table we got during the bulk download step
that we have five metadata files. Two of these should be the individual
and biospecimen files, and three of them are assay metadata files.
```{r}
download_table %>% 
  dplyr::select(name, metadataType, assay)
```

We are only interested in RNAseq data, so we will only read in the
individual, biospecimen, and RNAseq assay metadata files.
```{r}
# counts matrix
counts <- read_tsv("../data/htseqcounts_5XFAD.txt", show_col_types = FALSE)

# individual metadata
ind_meta <- read_csv("../data/Jax.IU.Pitt_5XFAD_individual_metadata.csv", show_col_types = FALSE)

# biospecimen metadata
bio_meta <- read_csv("../data/Jax.IU.Pitt_5XFAD_biospecimen_metadata.csv", show_col_types = FALSE)

#assay metadata
rna_meta <- read_csv("../data/Jax.IU.Pitt_5XFAD_assay_RNAseq_metadata.csv", show_col_types = FALSE)
```

Let’s examine the data and metadata files a bit before we begin our
analyses.

**Counts data**
```{r}
# Calling a tibble object will print the first ten rows in a nice tidy output; doing the same for a base R dataframe will print the whole thing until it runs out of memory. If you want to inspect a large dataframe, use `head(df)`
counts
```

The data file has a column of ENSEMBL gene ids and then a bunch of
columns with count data, where the column headers correspond to the
specimenIDs. These specimenIDs should all be in the RNAseq assay
metadata file, so let’s check.
```{r}
# what does the RNAseq assay metadata look like?
rna_meta
```

```{r}
# are all the column headers from the counts matrix (except the first "gene_id" column) in the assay metadata?
all(colnames(counts[-1]) %in% rna_meta$specimenID)
```

**Assay metadata** 

The assay metadata contains information about how data
was generated on each sample in the assay. Each specimenID represents a
unique sample. We can use some tools from dplyr to explore the metadata.
```{r}
# how many unique specimens were sequenced?
n_distinct(rna_meta$specimenID)
```

```{r}
# were the samples all sequenced on the same platform?
distinct(rna_meta, platform)
```

```{r}
# were there multiple sequencing batches reported?
distinct(rna_meta, sequencingBatch) 
```

**Biospecimen metadata** 

The biospecimen metadata contains specimen-level
information, including organ and tissue the specimen was taken from, how
it was prepared, etc. Each specimenID is mapped to an individualID.
```{r}
# all specimens from the RNAseq assay metadata file should be in the biospecimen file
all(rna_meta$specimenID %in% bio_meta$specimenID)
```

```{r}
# but the biospecimen file also contains specimens from different assays
all(bio_meta$specimenID %in% rna_meta$specimenID)
```

**Individual metadata** 

The individual metadata contains information about
all the individuals in the study, represented by unique individualIDs.
For humans, this includes information on age, sex, race, diagnosis, etc.
For MODEL-AD mouse models, the individual metadata has information on
model genotypes, stock numbers, diet, and more.
```{r}
# all individualIDs in the biospecimen file should be in the individual file
all(bio_meta$individualID %in% ind_meta$individualID)
```

```{r}
# which model genotypes are in this study?
distinct(ind_meta, genotype)
```

**Joining metadata** 

We use the three-file structure for our metadata
because it allows us to store metadata for each study in a tidy format.
Every line in the assay and biospecimen files represents a unique
specimen, and every line in the individual file represents a unique
individual. This means the files can be easily joined by specimenID and
individualID to get all levels of metadata that apply to a particular
data file. We will use the left_join() function from the dplyr package,
and the %\>% operator from the magrittr package. If you are unfamiliar
with the pipe, think of it as a shorthand for “take this (the preceding
object) and do that (the subsequent command)”. See here (https://magrittr.tidyverse.org/) 
for more info on piping in R.
```{r}
# join all the rows in the assay metadata that have a match in the biospecimen metadata
joined_meta <- rna_meta %>% #start with the rnaseq assay metadata
  left_join(bio_meta, by = "specimenID") %>%  #join rows from biospecimen that match specimenID 
  left_join(ind_meta, by = "individualID") # join rows from individual that match individualID

joined_meta
```

We now have a very wide dataframe that contains all the available
metadata on each specimen in the RNAseq data from this study. This
procedure can be used to join the three types of metadata files for
every study in the AD Knowledge Portal, allowing you to filter
individuals and specimens as needed based on your analysis criteria!
```{r}
library(lubridate)

# convert columns of strings to month-date-year format
joined_meta_time <- joined_meta %>% 
  mutate(dateBirth = mdy(dateBirth), dateDeath = mdy(dateDeath)) %>% 
  # create a new column that subtracts dateBirth from dateDeath in days, then divide by 30 to get months
  mutate(timepoint = as.numeric(difftime(dateDeath, dateBirth, units ="days"))/30) %>% 
  # convert numeric ages to timepoint categories
  mutate(timepoint = case_when(timepoint > 10 ~ "12 mo",
                               timepoint < 10 & timepoint > 5 ~ "6 mo",
                               timepoint < 5 ~ "4 mo"))

covars_5XFAD <- joined_meta_time %>%
  dplyr::select(individualID, specimenID, sex, genotype, timepoint) %>% distinct() %>% as.data.frame()
rownames(covars_5XFAD) <- covars_5XFAD$specimenID

head(covars_5XFAD)
```

We will save joined_meta for the next lesson.
```{r}
saveRDS(covars_5XFAD, file = "../data/covars_5XFAD.rds")
```

## Single Specimen files

For files that contain data from a single specimen (e.g. raw sequencing files, raw mass spectra, etc.), we can use the Synapse annotations to associate these files with the appropriate metadata.

Excercise 3: Use Explore Data to find all RNAseq files from the Jax.IU.Pitt_5XFAD study. If we filter for data where Study = “Jax.IU.Pitt_5XFAD” and Assay = “rnaSeq” we will get a list of 148 files, including raw fastqs and processed counts data.

Synapse entity annotations We can use the function synGetAnnotations to view the annotations associated with any file without downloading the file.
```{r}
# the synID of a random fastq file from this list
random_fastq <- "syn22108503"

# extract the annotations as a nested list
fastq_annotations <- synGetAnnotations(random_fastq)

fastq_annotations
```

The file annotations let us see which study the file is associated with (Jax.IU.Pitt.5XFAD), which species it’s from (Mouse), which assay generated the file (rnaSeq), and a whole bunch of other properties. Most importantly, single-specimen files are annotated with with the specimenID of the specimen in the file, and the individualID of the individual that specimen was taken from. We can use these annotations to link files to the rest of the metadata, including metadata that is not in annotations. This is especially helpful for human studies, as potentially identifying information like age, race, and diagnosis is not included in file annotations.
```{r}
# find records belonging to the individual this file maps to in our joined metadata
joined_meta %>% 
  filter(individualID == fastq_annotations$individualID[[1]])
```
## Annotations during bulk download

When bulk downloading many files, the best practice is to preserve the download manifest that is generated which lists all the files, their synIDs, and all their annotations. If using the Synapse R client, follow the instructions in the Bulk download files section above.

If we use the “Programmatic Options” tab in the AD Portal download menu to download all 148 rnaSeq files from the 5XFAD study, we would get a table query that looks like this:
```{r}
query <- synTableQuery("SELECT * FROM syn11346063.37 WHERE ( ( \"study\" HAS ( 'Jax.IU.Pitt_5XFAD' ) ) AND ( \"assay\" HAS ( 'rnaSeq' ) ) )")
```

As we saw previously, this downloads a csv file with the results of our AD Portal query. Opening that file lets us see which specimens are associated with which files:
```{r}
annotations_table <- read_csv(query$filepath, show_col_types = FALSE)

annotations_table
```

You could then use purrr::walk(download_table$id, ~synGet(.x, downloadLocation = )) to walk through the column of synIDs and download all 148 files. However, because these are large files, it might be preferable to use the Python client or command line client for increased speed.

Once you’ve downloaded all the files in the id column, you can link those files to their annotations by the name column.
```{r}
# We'll use the "random fastq" that we got annotations for earlier
# To avoid downloading the whole 3GB file, we'll use synGet with "downloadFile = FALSE" to get only the Synapse entity information, rather than the file. 
# If we downloaded the actual file, we could find it in the directory and search using the filename. Since we're just downloading the Synapse entity wrapper object, we'll use the file name listed in the object properties.

fastq <- synGet(random_fastq, downloadFile = FALSE)

# filter the annotations table to rows that match the fastq filename
annotations_table %>% 
  filter(name == fastq$properties$name)
```
## Multispecimen files

Multispecimen files in the AD Knowledge Portal are files that contain data or information from more than one specimen. They are not annotated with individualIDs or specimenIDs, since these files may contain numbers of specimens that exceed the annotation limits. These files are usually processed or summary data (gene counts, peptide quantifications, etc), and are always annotated with isMultiSpecimen = TRUE.

If we look at the processed data files in the table of 5XFAD RNAseq file annotations we just downloaded, we will see that it isMultiSpecimen = TRUE, but individualID and specimenID are blank:
```{r}
annotations_table %>% 
  filter(fileFormat == "txt") %>% 
  dplyr::select(name, individualID, specimenID, isMultiSpecimen)
```
The multispecimen file should contain a row or column of specimenIDs that correspond to the specimenIDs used in a study’s metadata, as we have seen with the 5XFAD counts file.
```{r}
# In this example, we take a slice of the counts data to reduce computation, transpose it so that each row represents a single specimen, and then join it to the joined metadata by the specimenID
counts %>% 
  slice_head(n = 5) %>% 
  t() %>% 
  as_tibble(rownames = "specimenID") %>% 
  left_join(joined_meta, by = "specimenID")
```

## Session Info
```{r}
sessionInfo()
```
